Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-03-31 curvature.tex
short note on correctness of the graph construction algorithm

\begin{theorem}
	\label{thm:construct_graph}
	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
\end{theorem}
\begin{proof}
	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g. red-black trees~\cite{guibas1978dichromatic}, as the vertex degrees are $\OhOf{n^2}$).
	As described above, the integer labels are simply the order of the input trees.
	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list, which can be stored as a fixed-size array.
	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time.
	Enumerating the neighbors of $T$ requires $\OhOf{n^3}$-time for each $T$, for a total of $\OhOf{mn^3}$-time.
	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
	We store the tree to index mappings for current vertices of $G$ in a trie~\cite{fredkin1960trie} using the $\OhOf{n}$-length Newick \cite{wiki:newick} representation of the tree as a string.
	These representations are made unique by ordering the tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
	This requires only $\OhOf{n}$-time for each tree (i.e. a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).
	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ $S$'s.
	Therefore the graph can be constructed in $\OhOf{mn^3}$-time.
\end{proof}
This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
The construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR graph from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
Moreover, although we do not study the 135,135-vertex 8-taxon rSPR graph in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus is infeasible for constructing larger rSPR tree graphs.
Thus, we believe this fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.

>>>>

\begin{theorem}
	\label{thm:construct_graph}
	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
\end{theorem}
\begin{proof}
	The correctness of the procedure follows by induction on the number of trees already processed, $i$, by observing that the procedure has constructed the subgraph of vertices $1, 2,  \ldots i$ and will construct the subgraph of vertices $1, 2, \ldots i+1$.

	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g. red-black trees~\cite{guibas1978dichromatic}, as the vertex degrees are $\OhOf{n^2}$).
	As described above, the integer labels are simply the order of the input trees.
	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list, which can be stored as a fixed-size array.
	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time.
	Enumerating the neighbors of $T$ requires $\OhOf{n^3}$-time for each $T$, for a total of $\OhOf{mn^3}$-time.
	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
	We store the tree to index mappings for current vertices of $G$ in a trie~\cite{fredkin1960trie} using the $\OhOf{n}$-length Newick \cite{wiki:newick} representation of the tree as a string.
	These representations are made unique by ordering the tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
	This requires only $\OhOf{n}$-time for each tree (i.e. a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).
	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ $S$'s.
	Therefore the graph can be constructed in $\OhOf{mn^3}$-time.
\end{proof}
This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
The construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR graph from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
Moreover, although we do not study the 135,135-vertex 8-taxon rSPR graph in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus is infeasible for constructing larger rSPR tree graphs.
Thus, we believe this fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.
